Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-15 appendix.tex
addressing comments

Consider the InvFels tree with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM This looks much better now. Can we get a subscript on tau for these calculations?
Define
\[
g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \{x_1,y_1,x_2,y_2,w\}),
\]
i.e., the generating probability of the $j$th site split as a function of $x_1$.
Recall the divergence
\[
\shannonDivergence_{\tau^*,t^*}(\tau,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, \tilde{t}),
\]
which, simplifying notation for clarity, we define as a function of $x_1$ as
\[
D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
\]
where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
By straightforward calculus,
\[
D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
\[
D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
\]
Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
\[
g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
\]
\[
g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
\]
and
\[
c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
\]
We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
This implies the above second derivative is negative.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau^*,t^*}(\tau, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \tilde{t}).
\]
As above, define the partial log-likelihood as a function of $x_1$ as
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
\]
where
\[
h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, \{x_1,y_1,x_2,y_2,w\}).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
%EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$, thus, $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

>>>>

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$.
We now show that, in the case of the InvFels tree, exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the likelihood, and that constraining both the top two branch parameters to be equal and the bottom two branch parameters to be equal during inference obtains the same maximum likelihood estimate as in the case of arbitrary branch parameters.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term (see Table~\ref{tab:gen-sitepatprob}).
For example, for site split $\{1\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{1\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    \Pr(\siteSplitRV=\{3\}\mid \tau_1, \tilde{t}) & = \frac{1}{8} (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
\begin{align*}
    \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{1\}$ and
\begin{align*}
        \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2); \\
    \Pr(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2); \\
    \Pr(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{3\}, \tau_1, \tilde{t}) &= \frac{1}{32}(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2);
\end{align*}
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM This looks much better now. Can we get a subscript on tau for these calculations?
%das: I changed all of them that were already subscripts, but did you also want subscripts on ones that didn't have them, e.g., D'' and g_1, etc.?
Define
\[
g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,w\}),
\]
i.e., the generating probability of the $j$th site split as a function of $x_1$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}),
\]
which, simplifying notation for clarity, we define as a function of $x_1$ as
\[
D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
\]
where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
By straightforward calculus,
\[
D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
\[
D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
\]
Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
\[
g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
\]
\[
g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
\]
and
\[
c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
\]
We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
This implies the above second derivative is negative.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
As above, define the partial log-likelihood as a function of $x_1$ as
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
\]
where
\[
h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,w\}).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
%EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
%das: added a sentence
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Thus, with $D''(x_1) < 0$, the sum $D''(x_1) + \tilde{\ell}''(x_1) < 0$, implying $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
